---
title: "Reproducible Research - Course Project 2"
author: "Numair Shahid Jamal Shaikh"
date: "29/07/2020"
output: html_document
keep_md: true
---


## Exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database - tracking characteristics of major storms and weather events in the United States - estimates of any fatalities, injuries, and property damage.

### Synopsis
This report is part of the second course project of the course - Reproducible Research offered by John Hopkins university under the Data Science Specialization on Coursera.
Every Year, storms and other natural calamities result in tremendous loss of life and property in the United States.
In this report we will analyze the data provided to us by the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database.
Analysis will include the number of injuries and fatalities as well as damage done to property and crops across the country due to these natural calamities.
While analyzing data we come across conclusions that factors such as tornadoes, floods and heat cause the most fatalities/injuries while drought, floods, hurricanes resulted in billions of dollars worth of damage to property and agricultural crops.

### **Problem Statement / Assignment (Task to be accomplished)**
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis.

*Questions :*

1. Across the United States, which types of events are most harmful with    respect to population health?
2. Across the United States, which types of events have the greatest        economic consequences?

### **Processing data.**

### *Load and setup the suitable libraries into R*
```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
```
### *Load the Data*
Set the working directory  to preferred location.
```{r}
setwd("C:/Users/Numair Shaikh/Desktop/R-Coursera")
if(!"data" %in% ls()) {
  data <- read.csv(bzfile("repdata_data_StormData.csv.bz2"), header=TRUE)
}
```

### *Basic Analysis* - Examining the dataset
Run the very useful function *str* on the dataset to know some vital information.
```{r}
str(data)
```
Examine the first few Entries in the data.
```{r}
head(data, 3)
```
Let's look at the variable names in the data.
```{r}
names(data)
```
We will refrain from looking the last few entries of the dataset at this stage due to the enormous remarks provided. 
```{r}
##tail(data)
```

### *Subset the required data*
We will only consider only a few variables in the data set and select variables which are relevant to the problem.

Variable(s) description :

1. BGN_DATE         : Beginning date of the event. 
2. EVTYPE           : The type of event/calamity, for example *Tornado,                        hurricane etc.*
3. FATALITIES       : The number of fatalities caused by the calamity.
4. INJURIES         : The number of injuries caused by the calamity.
5. PROPDMG          : Damage done to property.
6. PROPDMGEXP       : Scale of the damage done to property for example *K                       - thousands, M - Millions, B - Billions*.
7. CROPDMG          : Damage done to agricultural crops.
8. CROPDMGEXP       : Scale of the damage done to agricultural crops.
Let's also look at the first and the last 3 entries in the data.
```{r}
storm <- data[, c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
head(storm,3)
tail(storm,3)
```
By looking at the entries we can see that there is considerably less amounts of data recorded and made available in the 50s, 60s and 70s. This data may not be very useful to us, so lets separate out data ans store it in a new variable. This will also require us to clean one of the variable to separate out data based on the year in which the entry(ies) were recorded.
```{r}
storm$year <- as.Date(storm$BGN_DATE, format = "%m/%d/%Y %H:%M:%S")
storm$year <- year(storm$year)
table(storm$year)
storm <- subset(storm, storm$year>1970)
```

### **Cleaning the data**
Let us first work on damage done to Crops and property and calculate the total damage done by different calamities respectively.
```{r}
storm.damage <- storm[, c("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
```
We will convert the EXP variables so as to enable us to calculate the total loss in numeric value (US Dollars)
Initialize empty rows to Zero.
```{r}
storm.damage$CROPDMGEXP[!grepl("K|M|B",storm.damage$CROPDMGEXP)] <- 0
storm.damage$PROPDMGEXP[!grepl("K|M|B",storm.damage$PROPDMGEXP)] <- 0
```
Replace 'K' with 1000, 'M' with 1000000 and 'B' with 1000000000
```{r}
storm.damage$PROPDMGEXP[grepl("K",storm.damage$PROPDMGEXP)] <- 1000
storm.damage$PROPDMGEXP[grepl("M",storm.damage$PROPDMGEXP)] <- 1000000
storm.damage$PROPDMGEXP[grepl("B",storm.damage$PROPDMGEXP)] <- 1000000000
storm.damage$CROPDMGEXP[grepl("K",storm.damage$CROPDMGEXP)] <- 1000
storm.damage$CROPDMGEXP[grepl("M",storm.damage$CROPDMGEXP)] <- 1000000
storm.damage$CROPDMGEXP[grepl("M",storm.damage$CROPDMGEXP)] <- 1000000000
```
Multiply the values by scale and store the answer in separate variables.
```{r}
storm.damage$totalPropDmg <- storm.damage$PROPDMG*as.numeric(storm.damage$PROPDMGEXP)
storm.damage$totalCropDmg <- storm.damage$CROPDMG*as.numeric(storm.damage$CROPDMGEXP)
```
Calculate the cumulative damage done by a particular calamity.
Also, we merge the data set with respect to the variable "EVTYPE".
```{r}
dmgaggr <- aggregate(totalPropDmg~EVTYPE, data=storm.damage, FUN=sum)
dmgaggr1 <- aggregate(totalCropDmg~EVTYPE, data=storm.damage, FUN=sum)
dmgaggr <- merge(dmgaggr, dmgaggr1, by.y="EVTYPE")
```
Calculate the cumulative injuries and fatalities with respect to the calamities.
Also, we merge the data set with respect to the variable "EVTYPE".
```{r}
aggr <- aggregate(FATALITIES~EVTYPE, data=storm,FUN=sum)
aggr1 <- aggregate(INJURIES~EVTYPE, data=storm, FUN=sum)
aggr <- merge(aggr, aggr1, by.y="EVTYPE")
```
Merge the two datasets, ie; damages and Fatalities/Injuries
```{r}
storm.final <- merge(aggr, dmgaggr, by.y="EVTYPE")
```
We will group the data like STORM, HEAT, FLOOD etc by searching for keywords related to these calamities, also label and group all other data with the keyword "OTHER".
```{r}
storm.final$TYPE <- "OTHER"
storm.final$TYPE[grepl("AVALANC", storm.final$EVTYPE, ignore.case = TRUE)] <- "AVALANCHE"
storm.final$TYPE[grepl("BLIZZARD", storm.final$EVTYPE, ignore.case = TRUE)] <- "BLIZZARD"
storm.final$TYPE[grepl("SNOW|COLD|FREEZE|ICE", storm.final$EVTYPE, ignore.case = TRUE)] <- "SNOW"
storm.final$TYPE[grepl("FOG", storm.final$EVTYPE, ignore.case = TRUE)] <- "FOG"
storm.final$TYPE[grepl("STORM", storm.final$EVTYPE, ignore.case = TRUE)] <- "STORM"
storm.final$TYPE[grepl("HEAT", storm.final$EVTYPE, ignore.case = TRUE)] <- "HEAT"
storm.final$TYPE[grepl("FLOOD", storm.final$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
storm.final$TYPE[grepl("TORNADO", storm.final$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
storm.final$TYPE[grepl("CURRENT", storm.final$EVTYPE, ignore.case = TRUE)] <- "CURRENTS"
storm.final$TYPE[grepl("SLIDE", storm.final$EVTYPE, ignore.case = TRUE)] <- "LANDSLIDE"
storm.final$TYPE[grepl("HURRICANE", storm.final$EVTYPE, ignore.case = TRUE)] <- "HURRICANE"
storm.final$TYPE[grepl("LIGHTNING", storm.final$EVTYPE, ignore.case = TRUE)] <- "LIGHTNING"
```
Finally, We omit the rows containing NAs (if any) from the data.
```{r}
storm.final <- na.omit(storm.final)
```

### **Data Analysis**
*Aggregate the health variables (INJURIES & FATALITIES)*
```{r}
fatalities <- aggregate(FATALITIES~TYPE, data=storm.final, FUN=sum)
injuries <- aggregate(INJURIES~TYPE, data=storm.final, FUN=sum)
```

*Aggregate the Economic Variables*
```{r}
CropDmg <- aggregate(totalCropDmg~TYPE, data=storm.final, FUN=sum)
PropDmg <- aggregate(totalPropDmg~TYPE, data=storm.final, FUN=sum)
TotalDmg <- CropDmg
TotalDmg$totalCropDmg <- CropDmg$totalCropDmg+PropDmg$totalPropDmg
names(TotalDmg) <- c("EVENT", "DAMAGE")
```

### **Results**
*Across the United States, which types of events are most harmful with respect to population health?*
```{r}
ggplot(fatalities, aes(TYPE, FATALITIES, label=FATALITIES)) + geom_bar(stat="identity") + coord_flip() + xlab("Event") + ylab("Fatalities") + ggtitle("Fatalities caused due to different weather events")
```

We can conclude that Tornadoes and Heat related events are the most fatal weather events across the United States.
```{r}
ggplot(injuries, aes(TYPE, INJURIES, label=INJURIES)) + geom_bar(stat="identity") + coord_flip() + xlab("Event") + ylab("Injuries") + ggtitle("Injuries caused due to different weather events")
```

We can conclude from the above plot that tornadoes cause the most injuries as compared to other weather events.

*Across the United States, which types of events have the greatest economic consequences?*
```{r}
ggplot(TotalDmg, aes(EVENT, DAMAGE, label=DAMAGE)) + geom_bar(stat="identity") + coord_flip() + xlab("Event") + ylab("Damage in US Dollars") + ggtitle("Economic damage caused due to different weather events")
```

We can conclude that flood cause the most Economic damage than other weather events.

